---
title: "Enclosure Analysis"
format: html
editor: visual
---

# Analysis info

Looking at differences in mean soybean height between plants protected from browsing to those unprotected from browsing. Individuals exposed to browsing were exposed with variable treatment concentrations (0:control or 1:tannins). Additionally we investigated differences in mean height over time incorporating WAT (Week after application) as an additional predictor variable. Plot and Field as were included as nested random effects.

Response Variable:

-   Mean Height (cm)

Predictor Variables:

-   Treatment (0: Control, 1: Tannins)
-   WAT (Week After Application: 1-4)
-   Enclosure (0: Not Enclosed, 1: Enclosed)
-   Plot (1-20)
-   Field (1-20)

Below is our process of fitting our data to the best possible model.

# Initial Enclosure Height Analysis

```{r}
#| echo: true
#| include: false
library(tidyverse)
library(ggplot2)
library(glmmTMB)
```

## Reading in Data Source

```{r}
# Read the data
data <- read.csv("Data.csv", header=TRUE)
```

## Incorporating binary variable to account for added plants after week 2

```{r}
data$num_plants_change <- ifelse(data$WAT >= 2, 1, 0)  
```

## Sub setting data to:

**1) Convert predictor variables to factors with "levels".**

**2) Remove unneeded variables**

**3) Remove rows with missing Height values (missing values = dead plant or sampling error).**

**4) Only include measurements from weeks 1 through 4**

```{r}
# Convert categorical variables to factors
data$Field <- as.factor(data$Field)
data$Plot <- as.factor(data$Plot)
data$Treatment <- as.factor(data$Treatment)
data$Enclosure <- as.factor(data$Enclosure)
data$WAT <- as.factor(data$WAT)

# Remove rows with missing Height values and unused columns
data <- data |>
  filter(!is.na(Height)) |>
  select(-PercentDamage)

# Subset to remove WAT values at week zero
data <- data[data$WAT != "0", ]
```

## Exploratory Data Analysis

**Calculating and Displaying mean soybean height for treatments across Weeks after application.**

```{r}
# Calculate mean Height for each Treatment and WAT combination
summary_data <- data |>
  group_by(Enclosure,Treatment, WAT) |>
  summarise(mean_height = mean(Height, na.rm = TRUE))

# Displaying mean soybean height for treatments across Weeks after application
print(summary_data)
```

## Calculating standard error for each group

```{r}
# Calculate mean and SE for each group
enclosure_data <- data |>
  group_by(WAT, Enclosure, Treatment) |>
  summarise(
    mean_height = mean(Height),
    se = sd(Height) / sqrt(n()),  # Calculate SD from raw data (Height)
    .groups = "drop"
  )
```

# Plotting differences in Protected and Unprotected Plant Heights by Treatment and WAT

```{r}
# Create the plot with updated facet labels and error bars
ggplot(enclosure_data, aes(x = WAT, y = mean_height, fill = Enclosure)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_height - se, ymax = mean_height + se),
    position = position_dodge(width = 0.7),
    width = 0.2
  ) +
  facet_wrap(~factor(Treatment, levels = c(0, 1), labels = c("Tannin Apllication: 0%", "Tannin Application: 10%"))) +  # Modify facet labels
  labs(
    x = "Weeks After Tannin Application",
    y = "Mean Soybean Height (cm)",
    fill = "Protection from Browsing:"
  ) +
  scale_fill_discrete(labels = c("Unprotected", "Protected")) +
  scale_color_discrete(labels = c("No Repellent", "Repellent Applied")) +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 12),
    axis.text.x = element_text(size = 12)  # Adjust x-axis text size if necessary
  )

```

## Simple enclosure height analysis

**Fitting models for simple linear Regression**

```{r}
# Linear Regression for Height

lm_m_height_1 <- lm(formula = Height ~ Treatment, data = data)

lm_m_height_2 <- lm(formula = Height ~ Treatment * WAT, data = data)

lm_m_height_3 <- lm(formula = Height ~ Treatment * WAT + Enclosure, data = data)

lm_m_height_4 <- lm(formula = Height ~ Treatment + WAT + Enclosure + Plot, data = data) 
```

**Choosing the best model**

```{r}
#ANOVA for differences in models 
anova(lm_m_height_1, lm_m_height_2, lm_m_height_3, lm_m_height_4)

# AIC to pick the best model
AIC(lm_m_height_1, lm_m_height_2, lm_m_height_3, lm_m_height_4)
```

**Model lm_m_height_4 has the lowest AIC score.**

```{r}
summary(lm_m_height_4)
```

**Linear regression model indicates that treatment had no significant effect on soybean height. Model doesn't quite show differences height based on treatment and enclosure. An ANOVA model may better show the relationship between treatment and heights of plants unprotected vs those that were protected.**

### Mixed Effect Model with nested random effects incorparating added binary variable

```{r}
mem_encl <- glmmTMB(log(Height) ~ Treatment * Enclosure * WAT + (1 | Plot) + (1 | Field), data = data)
```

### Model Diagnostics

```{r}
AIC(mem_encl)
```

AIC with of model with binary variable is slightly lower then without (\~2 points lower)

```{r}
summary(mem_encl)
```

## Model Diagnostics

### Extracting model residuals

```{r}
residuals <- resid(mem_encl)
fitted <- fitted(mem_encl)

# Plot residuals vs fitted values
plot(fitted, residuals, xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")
```

### Normality in Residuals

```{r}
qqnorm(residuals)
qqline(residuals, col = "red")
```

## 3-way anova

```{r}
# Main effect of Enclosure on height only
# Plants within enclosures were not treated with tannins
anova_height_encl <- aov(Height ~ Treatment * WAT + Enclosure, data = data)
```

```{r}
AIC(anova_height_encl)
```

```{r}
summary(anova_height_encl)
```

## 3-way anova with Log Height

```{r}
# Main effect of Enclosure on height only
# Plants within enclosures were not treated with tannins
anova_height_encl_log <- aov(log(Height) ~ Treatment * WAT + Enclosure, data = data)

anova_height_encl_log_3way <- aov(log(Height) ~ Treatment * WAT * Enclosure, data = data)
```

```{r}
AIC(anova_height_encl_log)
AIC(anova_height_encl_log_3way)
```

AIC for model without 3-way interaction is slightly smaller

```{r}
# Chi-sq test to determine if I should keep 3-way interaction 
anova(anova_height_encl_log_3way, anova_height_encl_log, test = "Chisq")
```

p-value ≤ 0.05, keeping the interaction is justified

```{r}
summary(anova_height_encl_log_3way)
```

**ANOVA model summary**

Main Effects:

Treatment (p = 0.0257, F = 4.987): There is a statistically significant effect of tannin treatment on log-transformed soybean height. Plants treated with tannins (Treatment = 1) likely differed in height compared to the control (Treatment = 0).

WAT (Week After Application) (p \< 2e-16, F = 292.214): There is a highly significant effect of time on soybean height. Height changes significantly over the four weeks after application.

Enclosure (p \< 2e-16, F = 215.933): There is a very strong significant effect of enclosure, meaning plants protected from browsing were significantly different in height compared to those exposed.

Two-Way Interactions:

Treatment × WAT (p = 0.9041, F = 0.189): No significant interaction. The effect of tannin treatment on height does not change significantly over time.

Treatment × Enclosure (p = 0.0298, F = 4.731): Significant interaction. The effect of tannin treatment on soybean height depends on whether the plants were enclosed or exposed to browsing.

WAT × Enclosure (p = 0.0127, F = 3.624): Significant interaction. The effect of time on soybean height differs between enclosed and non-enclosed plants.

Three-Way Interaction:

Treatment × WAT × Enclosure (p = 0.9463, F = 0.123): Not significant. This suggests that the combined effects of treatment, time, and enclosure on height are not meaningfully different from the additive effects of these variables.

Residuals: The residual mean square is 0.092, which reflects the unexplained variance in log-transformed height.

## Continuing analysis for other response variables

### [Click here for the analysis on Browsing Damage](Percent_Damage.html)
