---
title: "yield"
format: html
editor: visual
---

# Analysis info

Looking at differences in yield (bushels/acre) based on tannin concentration (0%:control or 10%:treatment).

# Initial Yield Analysis

## Loading Data

```{r}
#| echo: true
#| include: false
library(tidyverse)
library(glmmTMB)
library(ggplot2)
```

### Reading in Data Source

```{r}
# Read the data
yield_data <- read.csv("yield_data.csv", header=TRUE)
```

```{r}
yield_data
```

# Creating a plot

```{r}
ggplot(yield_data, aes(x = Tannin.Concentration, y = Yield, fill = Tannin.Concentration)) +
  geom_boxplot() +
  labs(
    x = "Tannin Concentration",
    y = "Yield (bushels/acre)"
  ) +
  theme(
    legend.position = "top",  # Remove the legend
    plot.margin = margin(t = 20, b = 20, l = 20, r = 20),  # Add space between the figure and the edges
    strip.text = element_text(size = 12),
    axis.text.x = element_text(size = 12)
  )

```

## Simple Linear Regression

```{r}
# Linear Regression for Yield (Bushels/Acre)

mem_yield_1 <- glmmTMB(Yield ~ Tannin.Concentration, data = yield_data)

mem_yield_2 <- glmmTMB(Yield ~ Tannin.Concentration + Plot, data = yield_data) # Plot main effect

mem_yield_3 <- glmmTMB(Yield ~ Tannin.Concentration + (1 | Plot), data = yield_data) # Plot random effect
```

## Model Comparison using ANOVA

```{r}
# Compare models using likelihood ratio test
anova(mem_yield_1, mem_yield_2, mem_yield_3)
```

## Model comparison using AIC

```{r}
AIC(mem_yield_1, mem_yield_2, mem_yield_3)
```

## Summary of best Model

```{r}
summary(lm_yield_2)
```

**Model indicates that differences in yield based on tannin concentrations are insignificant.**

**Would an ANOVA be a more appropriate model?**
